SU($n$) spin models near the Heisenberg critical point

### 2020-03-19 collective_sun/sun_perturbation.tex
overhaul of main text and notation in the multi-body notes these notes are now structures to consider general M-body perturbations from the start

\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}

\newcommand{\DD}{\mathbb{D}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

\DeclareMathOperator{\diag}{diag}
\newcommand{\ul}{\underline}

@@

\title{Perturbing SU($n$)-symmetric interactions}%
\author{Michael A. Perlin}%
\date{\today}

@@

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\mu,\nu$ index states in an
orthonormal basis for the $n$-dimensional Hilbert space $\H_n$ of a
single spin; $h_{pq}$ are scalar coefficients; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is a
multilevel generalization of the SU(2)-symmetric spin interaction
$\v s_p\c\v s_q$ for $\v s=\p{s_\x,s_\y,s_\z}$, with
$s_\alpha=\sigma_\alpha/2$ and $\sigma_\alpha$ a Pauli matrix for an
individual spin.

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_X \sum_p v_{Xp} X_p = \sum_X \v v_X\c\v X,
  &&
  \V_2 \equiv \sum_O \sum_{p<q} w_{Opq} O_{pq}
  = \sum_O \v{\m w}_O\c \v{\m O}
  \label{eq:perturbations}
\end{align}
where $p,q$ index individual spins; $v_{Xp}$ and $w_{Opq}$ are real
numbers (coefficients); $X$ is a trace-zero single-spin operator on
$\H_n$; $O$ is a trace-zero two-spin operator on $\H_n\otimes\H_n$
that obeys pair-wise permutational symmetry,
i.e.~$O_{pq} = \Pi_{pq} O_{pq} \Pi_{pq} = O_{qp}$; $\v v_X$ and $\v X$
are $N$-component vectors defined by
\begin{align}
  \v v_X \equiv \sum_p v_{Xp} \ket{p},
  &&
  \v X \equiv \sum_p X_p \ket{p};
\end{align}
and finally $\v{\m w_O}$ and $\v{\m O}$ are ${N \choose 2}$-component
vectors defined by
\begin{align}
  \v{\m w}_O \equiv \sum_{p<q} w_{Opq} \ket{\set{p,q}},
  &&
  \v{\m O} \equiv \sum_{p<q} O_{pq} \ket{\set{p,q}},
\end{align}
where the basis vectors $\ket{\set{p,q}}$ are labeled by a choice of
two distinct spins $p,q$.

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of permutationally symmetric states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{pq}$.  We will adopt the restriction that all $h_{pq}<0$
throughout these notes, but note that this restriction can be relaxed
to the assumption that the initial state of all spins is
permutationally symmetric (e.g.~a spin-polarized state).  We will also
assume throughout these notes that the permutationally symmetric
manifold $\M_0$ is gapped away from all other states by an interaction
energy difference that is large compared to any coupling between
$\M_0$ and its orthogonal complement.  Power-law couplings of the form
$h_{pq}=-h/\abs{p-q}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap when $\alpha\le D$
(see Section \ref{sec:spectral_gap}).

The effective Hamiltonian $H_M$ induced on the ground-state manifold
$\M_0$ by an $M$-body perturbation $\V_M$ through second order in
perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_M = H_M^{(1)} + H_M^{(2)},
  &&
  H_M^{(1)} = \P_0 \V_M \P_0,
  &&
  H_M^{(2)} = - \P_0 \V_M \E \V_M \P_0,
  &&
  \E \equiv \sum_{\Delta\ne0} \f{\P_\Delta}{\Delta},
  \label{eq:H_M}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$ with interaction energy $\Delta$ above
that of permutationally symmetric manifold $\M_0$; that is, the
interaction energy of states in the support of $\P_\Delta$ is
$E_0+\Delta$, where
\begin{align}
  E_0 = \sum_{p<q} h_{pq}
\end{align}
is the interaction energy of permutationally symmetric states
$\ket\psi\in\M_0$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-body perturbations}
\label{sec:single_body_pert}

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQPS \sum_{X,p} v_{Xp} X_1
  \EQPS \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \sum_p X_p,
\end{align}
are respectively the mean value in $\v v_X$ and the collective version
of $X$.  In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \sum_p \p{\sum_q h_{pq} v_{Xq} - h_p v_{Xp}} X_p
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ (indexed by $\Delta$) is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  \label{eq:V1_eig}
\end{align}
where $\m h\equiv\sum_{p,q}h_{pq}\op{p}{q}$ is a matrix of all
couplings $h_{pq}$; the vector $\v h\equiv\sum_{p,q}h_{pq}\ket{p}$ is
the sum of all columns of $\m h$; and $\diag\v h$ is a matrix with
$\v h$ on the diagonal and zeroes everywhere else.  Solving the
general eigenvalue problem in \eqref{eq:V1_eig} requires diagonalizing
the $N\times N$ matrix $\m h-\diag\v h$.  When the interaction
Hamiltonian $H_0$ is translationally invariant, however, this
eigenvalue problem can be solved analytically; we provide the
corresponding solution at the end of Section \ref{sec:H_V1_psi}.

If we {\it construct} a perturbation $\V_1^\Delta$ with coefficients
$\v v_X^\Delta$ that satisfy \eqref{eq:V1_eig} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_1^\Delta \ket\psi = \p{E_0 + \Delta} \V_1^\Delta \ket\psi.
\end{align}
Interestingly, the energy of the state $\V_1^\Delta\ket\psi$ depends
only on the coefficients $\v v_X^\Delta$, and is entirely independent
of the state $\ket\psi$ or the choice of a trace-zero single-spin
operator $X$ used to build $\V_1^\Delta$.  Finding operators
$\V_1^\Delta$ that generate eigenvectors of the interaction
Hamiltonian $H_0$ when they act on permutationally symmetric states
$\ket\psi\in\M_0$ thus reduces to finding eigenvectors of the
interaction matrix $\m h-\diag\v h$.

Any coefficient vector $\v v_X$ can be expanded into its projections
$\v v_X^\Delta$ onto the eigenspace of vectors satisfying
\eqref{eq:V1_eig} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the permutationally
symmetric manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Section \ref{sec:PXYP}.  In total, we find that
\begin{align}
  H_1^{(2)}
  \EQPS \f1{N\p{N-1}} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:H_1_2}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$; and $\v v_X^\Delta$ is the
projection of $\v v_X$ onto the $\Delta$-eigenspace of the matrix
$\m h-\diag\v h$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two-body perturbations}
\label{sec:two_body_pert}

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
    - \tilde h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \tilde h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h}_2 \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}_2
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \tilde h_{pq} \bra{\set{p,q}}},
  &&
  \v{\m w} \equiv \sum_{p<q} w_{pq} \ket{\set{p,q}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} nominally requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}_2$.  When the interaction Hamiltonian $H_0$ and
perturbation $\V_M$ are translationally invariant, however, this
eigenvalue problem can be reduced to that of diagonalizing a matrix of
dimensions $\sim N\times N$, and isotropy can reduce this problem even
further.  We discuss these reductions in Section
\ref{sec:multi_body_symmetries}.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\Z}{\mathcal{Z}}

\newcommand{\PP}{\mathbb{P}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\diag}{diag}
\newcommand{\ul}{\underline}

@@

\title{SU($n$) spin models near the Heisenberg critical point}%
\author{Michael A. Perlin}%
\date{\today}

@@

\red{TODO: write an introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The SU($n$) Heisenberg model}

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{\p{p,q}\in\C_N\p{2}} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu\in\ZZ_n} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\C_N\p{M}$ is the set of all
choices of $M$ spins out of $N$; $h_{pq}$ are scalar coupling
constants; $\mu,\nu$ index states in an orthonormal basis for the
$n$-dimensional Hilbert space of a single spin; $\ZZ_N$ is the set of
integers modulo $N$; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is
essentially a multilevel generalization of the SU(2)-symmetric spin
interaction $\v s_p\c\v s_q$, with $\v s\equiv\p{\v s_\x,s_\y,s_\z}$,
$s_\alpha\equiv\sigma_\alpha/2$, and $\sigma_\alpha$ a single-spin
Pauli operator.

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of permutationally symmetric (PS) states
that are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{pq}$.  We will assume that all $h_{pq}<0$ throughout these
notes, but comment that this restriction can be relaxed to the
assumption that the initial state of all spins is permutationally
symmetric (e.g.~a spin-polarized state).  We will also assume that the
PS manifold $\M_0$ is gapped away from all orthogonal states by an
interaction energy $\Delta_{\t{gap}}>0$.  Power-law couplings of the
form $h_{pq}\sim-1/\abs{p-q}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap $\Delta_{\t{gap}}$
for long-range interactions with $\alpha\le D$ (see Appendix
\ref{sec:spectral_gap}).

The PS manifold $\M_0$ is spanned by a basis of states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu$ of the single-spin state $\mu$, and $\sum_\mu m_\mu=N$.  In
the case of $n=2$, for example, the PS manifold $\M_0$ is precisely
the Dicke manifold\cite{dicke1954coherence} of $N$ qubits, spanned by
Dicke states $\set{\ket{m_0,m_1}}$ for which $m_0+m_1=N$ and the
integer $m_0$ ($m_1$) indicates the total occupation number of qubit
state $\ket{0}$ ($\ket{1}$).  The dimension of the PS manifold $\M_0$
is determined by the number of ways to assign $N$ spins to $n$ states,
which is ${N+n-1 \choose n-1}\sim N^{n-1}$.  The energy of any PS
state $\ket\psi\in\M_0$ with respect to the interaction Hamiltonian
$H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\p{p,q}\in\C_N\p{2}} h_{pq}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Breaking SU($n$) symmetry (perturbatively)}
\label{eq:pert_theory}

In order to induce non-trivial dynamics for initially PS states
$\ket\psi\in\M_0$, which are otherwise stationary eigenstates of the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0}, we will add to the
Hamiltonian an operator of the general form \red{[TODO: consider more
  general sums of such operators]}
\begin{align}
  X_{\m w} \equiv \sum_{k\in\C_N\p{M}} w_k X_k,
  \label{eq:pert}
\end{align}
where $X$ is an $M$-spin operator that is invariant under arbitrary
permutation of its tensor factors; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is
a choice of $M$ distinct spins that $X_k$ acts on; $\C_N\p{M}$ is the
set of all choices of $M$ elements from a set of size $N$; and $\m w$
is a dimension-$M$ tensor with scalar entries $w_k$.  Even though only
${N\choose M}$ entries of $\m w$ are used to define $X_{\m w}$ in
\eqref{eq:pert}, for later convenience we define $w_k$ for all indices
$k\in\ZZ_N^M$ by enforcing that $\m w$ is symmetric with respect to
any permutation of its indices, and that $w_k=0$ for any $k$ with
repeated values. \red{[TODO: relax the restrictions on $X$ and $\m w$
  to allow for more general perturbations, e.g.~of the form
  $s_\x\otimes s_\y-s_\y\otimes s_\x$]}

Single-body ($M=1$) operators correspond to classical external fields,
e.g.~$s_\z$ for a magnetic field, and two-body ($M=2$) operators
correspond to two-body interactions that deviate from $H_0$, such as
$s_\z\otimes s_\z$ interactions in an XXZ model\red{[TODO:cite]} near
the Heisenberg critical point.  While we would be hard-pressed to find
a physical implementation of multi-body spin interactions with $M>2$,
considering this case will be necessary to construct low-lying
eigenstates of $H_0$.  Constructing low-lying eigenstates of $H_0$
will in turn enable simulating the dynamics induced by
non-perturbative additions $X_{\m w}$ to the SU($n$) Heisenberg model,
with norms $\norm{X_{\m w}}\sim\Delta_{\t{gap}}$.

When $X_{\m w}$ is sufficiently small, we can treat its effect on the
PS manifold $\M_0$ perturbatively.  The effective Hamiltonian
$H_{\t{eff}}$ induced by $X_{\m w}$ on the ground-state manifold
$\M_0$ through second order in perturbation theory
is\cite{bravyi2011schrieffer, perlin2019effective}
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
  &&
  H_{\t{eff}}^{(1)} = \P_0 X_{\m w} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 X_{\m w} \P_\Delta X_{\m w} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ is straightforward to
simplify using the permutational symmetry of $\M_0$:
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \bar w\,\col{X},
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; $\bar w$
is an average of $w_k$ over spin choices $k\in\C_N\p{M}$; and
$\col{X}$ is the collective version of $X$, i.e.
\begin{align}
  \bar w \equiv { N \choose M }^{-1} \sum_{k\in\C_N\p{M}} w_k,
  &&
  \col{X} \equiv \sum_{k\in\C_N\p{M}} X_k,
\end{align}
An immediate consequence of \eqref{eq:H_eff_1} is the fact that
perturbing SU(2)-symmetric interactions by two-body $s_\z\otimes s_\z$
interactions yields the one-axis twisting (OAT)
Hamiltonian\cite{kitagawa1993squeezed, ma2011quantum}
$H_{\t{OAT}}=\chi S_\z^2$ at first order in perturbation theory, with
the OAT strength $\chi$ equal to the mean coefficient of the two-body
$s_\z\otimes s_\z$ terms.

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$ (with respect to $H_0$).  This projector essentially
picks off the part of $X_{\m w}$ that is strictly off-diagonal with
respect to the ground- and excited-state manifolds $\E_0$ and
$\E_\Delta$.  To decompose $X_{\m w}$ into components that generate
states of definite excitation energy when acting on PS states $\M_0$,
we pick an arbitrary state $\ket\psi\in\M_0$ and expand (see Appendix
\ref{sec:eigenstates})
\begin{align}
  H_0 X_{\m w} \ket\psi
  = E_0 X_{\m w} \ket\psi + X_{\hat{\m h}\p{\m w}} \ket\psi,
  \label{eq:diagnosis}
\end{align}
where $\hat{\m h}$ is a linear map on the space of dimension-$M$
tensors, determined entirely by the couplings $h_{pq}$ of the
interaction Hamiltonian $H_0$.  The image $\hat{\m h}\p{\m w}$ of
$\m w$ under $\hat{\m h}$ is then itself a dimension-$M$ tensor that
defines the $M$-body operator $X_{\hat{\m h}\p{\m w}}$.  To write out
$\hat{\m h}\p{\m w}$ explicitly, we first define a matrix
$\m h\equiv\sum_{p,q\in\ZZ_N} h_{pq} \op{p}{q}$ of all couplings in
$H_0$ by enforcing that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We denote a
contraction of $\m h$ with the $a$-th index of $\m w$ by
$\m h \circ_a\m w$, such that the $k$-th entry of $\m h \circ_a\m w$
is
\begin{align}
  \p{\m h \circ_a \m w}_k
  = \sum_{p\in\ZZ_N\setminus k} h_{k_a p} w_{k_{a:p}},
\end{align}
where $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of indices;
$k_{a:p}\equiv\p{k_1,k_2,\cdots,k_{a-1},p,k_{a+1},\cdots,k_M}$ is the
same list $k$, but with the $a$-th index $k_a$ replaced by $p$; and we
explicitly exclude terms with $p\in k$ that vanish due to the fact
that $\m h$ and $\m w$ are strictly off-diagonal.  We then define a
dimension-$M$ tensor $\tilde{\m h}$ with entries
\begin{align}
  \tilde h_k \equiv \sum_{\substack{p\in k\\q\in\ZZ_N\setminus k}} h_{pq}
\end{align}
for all $k\in\ZZ_N^M$, and we denote an element-wise (Kronecker)
product of $\tilde{\m h}$ and $\m w$ by $\tilde{\m h} * \m w$, with
entries $\tilde h_k w_k$.  These definitions allow us to expand
\begin{align}
  \hat{\m h}\p{\m w}
  \equiv \sum_{a\in\ZZ_M} \m h \circ_a \m w - \tilde{\m h} * \m w.
\end{align}
Viewed an operator on the vector space of dimension-$M$ tensors, the
map $\hat{\m h}$ is real and symmetric under transposition, and is
therefore diagonalizable.  Diagonalizing this operator yields
eigenvalues $\Delta$ and eigenvectors $\m m_\Delta$ for which
$\check{\m h}\p{\m m_\Delta}=\Delta\m m_\Delta$.  Finding eigenvalues
and eigenvectors of $\hat{\m h}$ nominally requires diagonalizing a
matrix of dimensions
${N\choose M}\times{N\choose M}\sim N^M\times N^M$, but symmetries of
$\hat{\m h}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem.  We
discuss this reduction in Appendix \ref{sec:symmetries}.

Defining $\m w_\Delta$ as the projection of the tensor $\m w$ onto the
$\Delta$-eigenspace of $\check{\m h}$, we can expand
\begin{align}
  \m w = \sum_\Delta \m w_\Delta,
  &&
  \hat{\m h}\p{\m w} = \sum_\Delta \Delta \m w_\Delta,
\end{align}
and in turn
\begin{align}
  X_{\m w} = \sum_\Delta X_{\m w_\Delta},
  &&
  X_{\hat{\m h}\p{\m w}} = \sum_\Delta \Delta X_{\m w_\Delta}.
\end{align}
Substituting these expansions into \eqref{eq:diagnosis}, we find
\begin{align}
  H_0 X_{\m w} \ket\psi
  = \sum_\Delta\p{E_0 + \Delta} X_{\m w_\Delta} \ket\psi,
\end{align}
which implies that the action of $X_{\m w_\Delta}$ on a PS state
$\ket\psi\in\M_0$ generates a state of definite interaction energy
$E_0+\Delta$.  It follows that
$\P_\Delta X_{\m w} \P_0 = \P_\Delta X_{\m w_\Delta} \P_0 = X_{\m
  w_\Delta} \P_0$, and so
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0}
  \f1\Delta \P_0 X_{\m w_\Delta} X_{\m w_\Delta} \P_0.
\end{align}
The product $\P_0 X_{\m w_\Delta} X_{\m w_\Delta} \P_0$ generally
includes $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator ($M=1$), then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\var\p{\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}},
  &&
  \var\p{\m m} \equiv \f1N \sum_{p\in\ZZ_N} \p{m_p-\bar m}^2,
\end{align}
which in the SU(2) case implies that an inhomogeneous magnetic field
($X\propto\sigma_\z$) yields an OAT Hamiltonian at second order
perturbation theory.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Beyond ground-state perturbation theory}
\label{eq:excited_states}

In order to find the effective Hamiltonian $H_{\t{eff}}^{(2)}$ induced
by the $M$-body operator $X_{\m w}$ on the PS manifold $\M_0$ at
second order in perturbation theory, we constructed states with
definite excitation energy $\Delta$ with respect to $H_0$.  These
states are generated by the action of an $M$-body operator
$X_{\m w_\Delta}$ on a PS state $\ket\psi\in\M_0$.  We denote the
manifold of all states that can be generated by the action of some
$M$-body operator, but not any $r$-body operator with $r<M$, by
$\M_M$; we will refer to $\M_M$ as the $M$-body excitation manifold.
We then denote the a direct sum of all $\M_r$ for $r\le M$ by
$\ul{\M}_M$, i.e.~$\ul{\M}_M\equiv\bigoplus_{r\le M}\M_r$.  The
manifold $\ul{\M}_M$ consists of all states accessible by the action
of $r$-body operators with $r\le M$ on PS states in $\M_0$.

By construction, the $M$-body operator $X_{\m w}$ generally couples PS
states in $\M_0$ to asymmetric states in $\ul{\M}_M\setminus\M_0$.  If
the coupling between $\M_0$ and its complement is weak compared to the
spectral gap $\Delta_{\t{gap}}$ of $H_0$, then we can treat the effect
of $X_{\m w}$ perturbatively and derive an effective Hamiltonian
$H_{\t{eff}}$ on $\M_0$.  This ground-state perturbation theory
formally breaks down when the operator norm
$\norm{X_{\m w}}\ge\Delta_{\t{gap}}/2$, in which case the operator
$X_{\m w}$ first couples $\M_0$ to $\ul{\M}_M$, then $\ul{\M}_M$ to
$\ul{\M}_{2M}$, and so on.  Even if
$\norm{X_{\m w}}>\Delta_{\t{gap}}/2$, however, the energetic penalty
incurred by $H_0$ from breaking the permutational symmetry of an
initial state $\ket\psi\in\M_0$ suppresses the leakage of population
into manifolds $\M_r$ of increasing $r$.  One might therefore capture
the dynamics of an initial PS state $\ket\psi\in\M_0$ by projecting
$X_{\m w}$ onto $\ul{\M}_{r_{\t{max}}}$ for some $r_{\t{max}}\ge M$,
resulting in a theory analogous to first-order perturbation theory on
$\ul{\M}_{r_{\t{max}}}$.  In the perturbative regime
$\norm{X_{\m w}}<\Delta_{\t{gap}}/2$, such a theory captures all
virtual processes through order $2\floor{r_{\t{max}}/M}+1$ in a formal
perturbation theory on $\M_0$.  In the non-perturbative regime
$\norm{X_{\m w}}\ge\Delta_{\t{gap}}/2$, the error incurred by
truncating the Hilbert space of a spin system past
$\ul{\M}_{r_{\t{max}}}$ is not clear a priori.  Nonetheless,
simulations restricted to $\ul{\M}_{r_{\t{max}}}$ can be benchmarked
by examining the maximum populations within all $\M_r$ for
$r\le r_{\t{max}}$ throughout simulation.  As long as these
populations fall off with increasing $r$ and the maximal population
within large-$r$ manifolds $\M_r$ is small, one should expect such
simulations to faithfully capture (at least qualitatively) the
dynamical behavior of the spin system under consideration.
